---
title: "Housing Prices Prediction"
author: "Scott Lovelady"
date: "12/5/2021"
output:
  html_document:
    df_print: paged
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
options(repos = list(CRAN="http://cran.rstudio.com/"))
```


## loading packages
```{r}
#check if packages are installed. if not, install the package
pckg=c("ggplot2","dplyr","ggmap","caret","leaps","ranger")
for(i in 1:length(pckg)) 
   {
      print(pckg[i])
      if (!is.element(pckg[i], installed.packages()[,1]))
      install.packages(pckg[i], dep = TRUE)
      require(pckg[i], character.only = TRUE)
}
```

## User created functions
```{r}
# NA value counter
na_counter <- function(dset) {
  missing_values <- which(colSums(is.na(dset)) > 0)
  sort(colSums(sapply(dset[missing_values], is.na)), decreasing = TRUE)
}

# Kaggle submission creator
submission <- function(model, test, method = 'standard') {
  if (method == 'log') { # if the target variable uses log in the model we need to convert this back for the prediction.
    a <- test %>% 
      select(Id) %>% 
      mutate(SalePrice = exp(predict(object = model, newdata = test)))
    write.csv(a,'submission_file.csv', row.names=FALSE)
  } 
  else {
    a <- test %>% 
      select(Id) %>% 
      mutate(SalePrice = predict(object = model, newdata = test))
    write.csv(a,'submission_file.csv', row.names=FALSE)
  }
}

rmse <- function(actual, predicted) sqrt(mean((actual - predicted)^2))

r2 <- function(actual, predicted){
  TSS <- sum((actual - mean(actual))^2)
  RSS <- sum((actual - predicted)^2)
  1 - RSS/TSS
}

# Calculates mode of a data set
calc_mode <- function(x){ # Function found: https://www.codingprof.com/how-to-replace-nas-with-the-mode-most-frequent-value-in-r/
  # List the distinct / unique values
  distinct_values <- unique(x)
  # Count the occurrence of each distinct value
  distinct_tabulate <- tabulate(match(x, distinct_values))
  # Return the value with the highest occurrence
  distinct_values[which.max(distinct_tabulate)]
}
```

## Importing data
```{r}
# Import data
training <- read.csv('train.csv')
test <- read.csv('test.csv')
```

## Understanding the data 
```{r}
# Glimpse of data
glimpse(training)

# Summary of data
summary(training)
```

## Understanding the data with visualizations
```{r}
# Plotting distribution of sale price
ggplot(training, aes(x = SalePrice)) +
  geom_histogram()+
  labs(title = "Distribution of Sales Prices")

# Plotting key data
ggplot(training, aes(x = SalePrice, y = LotArea)) +
  geom_point()+
  labs(title = "Sales Price by Lot Size")
```

# Data cleaning

## Combining datasets
```{r}
# creating identifiers so data sets can be separated after cleaning
training$ident <- T
test <- test %>%
  mutate(ident = F,
         SalePrice = NA)

# combining the data
df <- rbind(training, test)
summary(df)
```

## Missing data
```{r}
# Count missing values by column
na_counter(df)
```

## NA meaning none
```{r}
# Columns where NA means none
na_none <- c("Alley", "BsmtQual", "BsmtCond", "MasVnrType", "BsmtExposure", "BsmtFinType1", "BsmtFinType2", "FireplaceQu", "GarageType", "GarageFinish", "GarageQual", "GarageCond", "PoolQC", "Fence", "MiscFeature")

#assigns "None" to NAs for applicable columns identified above
df[na_none][is.na(df[na_none])] <- "None"

# Did it work?
na_counter(df)
```

## NA meaning 0
```{r}
# Columns where NA means 0
na_zero <- c("LotFrontage", "GarageYrBlt", "MasVnrArea", "BsmtFullBath", "BsmtHalfBath", "BsmtFinSF1", "BsmtFinSF2", "BsmtUnfSF", "GarageCars", "GarageArea")

#assigns 0 to NAs for applicable columns identified above
df[na_zero][is.na(df[na_zero])] <- 0

# Checking the number of missing values
na_counter(df)
```
## Dealing with the other NA's
```{r}
# Viewing the rows which have missing values to better decide what do do with them.
df_temp <- df %>%
          select(- SalePrice)

missing_values_rows <- df_temp[rowSums(is.na(df_temp)) > 0,]
missing_values_rows
```

```{r}
# Changing NA to most common value (mode)
df$MSZoning[is.na(df$MSZoning)] <- calc_mode(df$MSZoning)

# Changing NA to defined value
df$Utilities[is.na(df$Utilities)] <- "AllPub" # Default value
df$Electrical[is.na(df$Electrical)] <- "SBrkr" # Standard value
df$Functional[is.na(df$Functional)] <- "Typ" # From data description: (Assume typical unless deductions are warranted)
df$KitchenQual[is.na(df$KitchenQual)] <- "TA" # Typical / Average
df$SaleType[is.na(df$SaleType)] <- "Oth" # Other value
df$Exterior1st[is.na(df$Exterior1st)] <- "Other" # Other value
df$Exterior2nd[is.na(df$Exterior2nd)] <- "Other" # Other value

# Changing NA to calculation
df$TotalBsmtSF[is.na(df$TotalBsmtSF)] <- df$BsmtFinSF1 + df$BsmtFinSF2 + df$BsmtUnfSF #This equals 0, but it could be important if there were any other NA's in a new dataset

na_counter(df)
```
## Fixing wrong datatypes
```{r}
# Converting to factors
df <- df %>%
  mutate(MSSubClass = factor(MSSubClass),
         MSZoning = factor(MSZoning),
         Street = factor(Street, order = TRUE, levels = c('Grvl', 'Pave')),
         Alley = factor(Alley, order = TRUE, levels = c('None', 'Grvl', 'Pave')),
         LotShape = factor(LotShape, order = TRUE, levels = c('None', 'IR3', 'IR2', 'IR1', 'Reg')),
         LandContour = factor(LandContour),
         Utilities = factor(Utilities, order = TRUE, levels = c('ELO', 'NoSeWa', 'NoSewr', 'AllPub')),
         LotConfig = factor(LotConfig),
         LandSlope = factor(LandSlope, order = TRUE, levels = c('Sev', 'Mod', 'Gtl')),
         Neighborhood = factor(Neighborhood),
         Condition1 = factor(Condition1),
         Condition2 = factor(Condition2),
         BldgType = factor(BldgType),
         HouseStyle = factor(HouseStyle),
         RoofStyle = factor(RoofStyle),
         RoofMatl = factor(RoofMatl),
         Exterior1st = factor(Exterior1st),
         Exterior2nd = factor(Exterior2nd),
         MasVnrType = factor(MasVnrType),
         ExterQual = factor(ExterQual, order = TRUE, levels = c('None', 'Po','Fa','TA','Gd','Ex')),
         ExterCond = factor(ExterCond, order = TRUE, levels = c('None', 'Po','Fa','TA','Gd','Ex')),
         Foundation = factor(Foundation),
         BsmtQual = factor(BsmtQual, order = TRUE, levels = c('None', 'Po','Fa','TA','Gd','Ex')),
         BsmtCond = factor(BsmtCond, order = TRUE, levels = c('None', 'Po','Fa','TA','Gd','Ex')),
         BsmtExposure = factor(BsmtExposure, order = TRUE, levels = c('None', 'No','Mn','Av','Gd')),
         BsmtFinType1 = factor(BsmtFinType1, order = TRUE, levels = c('None', 'Unf','LwQ','Rec','BLQ','ALQ','GLQ')),
         BsmtFinType2 = factor(BsmtFinType2, order = TRUE, levels = c('None', 'Unf','LwQ','Rec','BLQ','ALQ','GLQ')),
         Heating = factor(Heating),
         HeatingQC = factor(HeatingQC, order = TRUE, levels = c('None', 'Po','Fa','TA','Gd','Ex')),
         CentralAir = factor(CentralAir, order = TRUE, levels = c('N','Y')),
         Electrical = factor(Electrical),
         KitchenQual = factor(KitchenQual, order = TRUE, levels = c('None', 'Po','Fa','TA','Gd','Ex')),
         Functional = factor(Functional, order = TRUE, levels = c('Sal','Sev','Maj2','Maj1','Mod','Min2','Min1','Typ')),
         FireplaceQu = factor(FireplaceQu, order = TRUE, levels = c('None', 'Po','Fa','TA','Gd','Ex')),
         GarageType = factor(GarageType),
         GarageFinish = factor(GarageFinish, order = TRUE, levels = c('None', 'Unf','RFn','Fin')),
         GarageQual = factor(GarageQual, order = TRUE, levels = c('None',  'Po','Fa','TA','Gd','Ex')),
         GarageCond = factor(GarageCond, order = TRUE, levels = c('None',  'Po','Fa','TA','Gd','Ex')),
         PavedDrive = factor(PavedDrive, order = TRUE, levels = c('None', 'N','P','Y')),
         PoolQC = factor(PoolQC, order = TRUE, levels = c('None', 'Fa','TA','Gd','Ex')),
         Fence = factor(Fence, order = TRUE, levels = c('None', 'MnWw','GdWo','MnPrv','GdPrv')),
         MiscFeature = factor(MiscFeature),
         MoSold = factor(MoSold),
         YrSold = factor(YrSold),
         SaleType = factor(SaleType),
         SaleCondition = factor(SaleCondition)) 
```

## Creating scores
```{r}
# Creating numerical scores from categorical data for quality variables
qual <- c('None' = 0, 'Po' = 1, 'Fa' = 2, 'TA' = 3, 'Gd' = 4, 'Ex' = 5)

df$ExterQualScore <- as.integer(recode(df$ExterQual, !!!qual))
df$ExterCondScore <- as.integer(recode(df$ExterCond, !!!qual))
df$BsmtQualScore <- as.integer(recode(df$BsmtQual, !!!qual))
df$BsmtCondScore <- as.integer(recode(df$BsmtCond, !!!qual))
df$HeatingQCScore <- as.integer(recode(df$HeatingQC, !!!qual))
df$KitchenQualScore <- as.integer(recode(df$KitchenQual, !!!qual))
df$FireplaceQuScore <- as.integer(recode(df$FireplaceQu, !!!qual))
df$GarageQualScore <- as.integer(recode(df$GarageQual, !!!qual))
df$GarageCondScore <- as.integer(recode(df$GarageCond, !!!qual))
df$PoolQualScore <- as.integer(recode(df$PoolQC, !!!qual))

# Creating aggregated scores 
df$TotalQualScore <- df$ExterQualScore + df$ExterCondScore + df$BsmtQualScore + df$BsmtCondScore + df$HeatingQCScore + df$KitchenQualScore + df$FireplaceQuScore + df$GarageQualScore + df$GarageCondScore + df$PoolQualScore

df$TotalExterQualScore <- df$ExterQualScore + df$ExterCondScore

df$TotalBsmtQualScore <- df$BsmtQualScore + df$BsmtCondScore

df$IndoorQualScore <- df$HeatingQCScore + df$KitchenQualScore + df$FireplaceQuScore

df$TotalGarageQualScore <- df$GarageQualScore + df$GarageCondScore 

# Total bathrooms
df$TotalBath <- df$BsmtFullBath + (df$BsmtHalfBath * 0.5) + df$FullBath + (df$HalfBath * 0.5)

# Total square feet
df$TotalSqFt <- df$TotalBsmtSF + df$GrLivArea

# Dropping columns that were aggregated and replaced above
df <- df %>%
  select(-ExterQualScore, -ExterCondScore, -BsmtQualScore, -BsmtCondScore, -HeatingQCScore, -KitchenQualScore, -FireplaceQuScore, -GarageQualScore, -GarageCondScore)
```
## Separating the training and the test data
```{r}
training <- df[df$ident==T,]
test  <- df[df$ident==F,]
```


# Model creation

## Single linear model
```{r}
# Baseline model
(house_price_model <- lm(SalePrice ~ Neighborhood, data = training)) %>%
  summary

submission(house_price_model, test, 0)
```


```{r}
#5 variable model
#iterate over training data to populate new data frame with column name and R squared of simple regressions for each column
h <- data.frame()
for(x in names(training)) {
    simple <- summary(lm(SalePrice ~ training[[x]], data = training))
    h <- rbind(h,list(x, simple$r.squared))
  } 
colnames(h) <- c("variable", "r2")

#remove duplicate rows, ordered by r2
h <- h %>%
  distinct() %>%
  filter(r2 < 1) %>%
  arrange(desc(r2))

#multiple linear model using top 5 columns (ordered by R^2) and skipping a column if it is likely multicollinear with a previously used column
(mmr2 <- lm(SalePrice ~ OverallQual + TotalSqFt + Neighborhood + GarageCars + TotalBath, data = training)) %>%
  summary

submission(mmr2, test, 'log')
```

```{r}
(model2 <- lm(log(SalePrice) ~ OverallQual + TotalSqFt + TotalQualScore + GarageCars + TotalBath, data = training)) %>%
  summary

submission(model2, test, 'log')
```
## Model performance with lm
Kaggle score: 0.16311 rank: #2933

```{r}
((model3 <- lm(log(SalePrice) ~ OverallQual + OverallQual^2 + Neighborhood + OverallCond + log(GrLivArea) + GrLivArea^4 + TotalBsmtSF * BsmtQual+ TotalBsmtSF^2 + MSZoning * log(LotArea) + YearBuilt + YearBuilt^20 + BldgType * HouseStyle + RoofMatl + KitchenQual + GarageArea + GarageType + IndoorQualScore + TotalBath + SaleCondition + LotConfig, data = training)) %>%
  summary)
```

## Model performance with random forest
```{r}
set.seed(123)

rf <- train(log(SalePrice) ~ OverallQual + OverallQual^2 + Neighborhood + OverallCond + log(GrLivArea) + GrLivArea^4 + TotalBsmtSF * BsmtQual+ TotalBsmtSF^2 + MSZoning + YearBuilt + YearBuilt^20 + BldgType * HouseStyle + RoofMatl + KitchenQual + GarageArea + GarageType + IndoorQualScore + TotalBath + SaleCondition + LotConfig, method = 'ranger',  data = training)
```

## RMSE on training dataset
```{r}
rmse(training$SalePrice, predict(rf, newdata= training)) 
```

## R-squared on training dataset
```{r}
r2(training$SalePrice, exp(predict(rf, newdata= training)))
```

#Testing the test data by calculating the Salesprice using the lm and random forest 
```{r}
submit <- test %>%
  mutate(SalePrice = (exp(predict(model3, newdata = test))+exp(predict(rf, newdata = test)))/2)
```

# RMSE and R-squared on test dataset
```{r}
rmse(submit$SalePrice, predict(rf, newdata= submit)) 

r2(submit$SalePrice, exp(predict(rf, newdata= submit)))
```

##Selecting ID and salesprice column
```{r}
submit <- submit %>%
  select(Id, SalePrice)
```

#Exporting the data to CSV file for submission 
```{r}
write.csv(submit,'stacked_submission_file.csv', row.names=FALSE)
```